Dynamic Walking and Whole-Body Motion Planning for Humanoid Robots: an Integrated Approach

### 2012-11-27 paper.tex
Add Acknowledgment section
Fix typos (preliminary pass).

In this paper, we present a planning algorithm that considers exact models of a humanoid
robot and its environment. It is used to solve navigation and manipulation problems. Our planner 
generates collision-free, statically stable 
geometric paths. We also present a constructive formal proof that these paths can be approximated by
collision-free, dynamically balanced walk trajectories. Our planner is designed for indoor
environments, where the floor is horizontal and flat. Because we do not compute explicitely footstep
positions at the planning stage, the presented planner is unable to plan motions in which the
robot steps over obstacles. This limitation is discussed in the paper.

@@

The problem of inverse kinematics for a humanoid robot, or any articulated
structure, is to compute a joint motion to achieve an end-effector pose. As the
robots we deal with are redundant, it is natural to take advantage of
this redundancy by specifying multiple tasks, potentially with
different priorities. This problem has been widely studied in robotics
planning and control literature, and many jacobian-based solutions have been
proposed, among which 
\cite{nakamura1986iks}, \cite{siciliano1991gfm},
\cite{baerlocher1998tpf} and \cite{khatib2004wbd}.
Obstacle avoidance can be taken into account with similar methods. To
do so, one has to include the obstacles as  constraints to
satisfy, see for example \cite{kanehiro2008lca}.
These methods are prone to fall into local minima, global motion 
planning is needed to overcome this limitation.
\cite{TouGieGoe2007} propose a motion generation method where tasks follow
trajectories define by cubic B-splines. The whole-body motion is optimized
with respect to the control point positions. This method can take into account
collision with simple obstacles.

@@

When  planning a  whole-body motion  for a  humanoid robot, one difficult
challenge is to cope with  the curse of dimensionality. The complexity
of   motion  planning  is   exponential  in   the  dimension   of  the
configuration  space ($\mathcal{C}$)  to explore.  When  dealing with
high-dimensional configuration  spaces, it is  typically impossible to
explicitly represent  them, leading to the use  of randomized sampling
techniques  to solve  global planning  problems. In  the  past fifteen
years,  \textit{Probabilistic Roadmaps} \cite{kavraki1996prp} and  
\textit{Rapidly exploring Random  Trees} (RRT) 
\cite{kuffner00rrtconnect}  have been  developed and  used to  solve many
high-dimensional   planning  problems, see \cite{Lav06} and \cite{choset2005prm} for comprehensive
overviews.
When  using  sampling  techniques  on  a humanoid  robot,  another difficulty
is to  take into  account stability  constraints,  i.e. to
generate  random  configurations   on  zero  volume  submanifolds  of
$\mathcal{C}$. This problem has been investigated with success during
the last few years, \cite{Berenson15032011} presents a exhaustive survey
of jacobian-based methods. Other recent contributions \cite{porta2012randomized}
present sophisticated constrained planning techniques based on higher-dimensional
continuation. Section \ref{sec:wb} presents a simple adaptation
of the RRT algorithm to constrained motion planning, that was first
introduced in \cite{dalibard09}.

@@

Another  field of  humanoid  robotics research  is  the generation  of
dynamically stable  walk patterns. Since  the introduction of  the ZMP
formalism  \cite{vukobratovic1969contribution},  several  methods  have  been  proposed  to
generate  walking  motions efficiently.   One  way  to  deal with  the
complexity  of a  humanoid  robotics  kinematic tree  is  to use  the
so-called "cart-table" simplified model \cite{kajita2003biped}. Based this
model,  planning a trajectory  for the  ZMP is  reduced to  planning a
trajectory  for  the Center  of  Mass (CoM)  of  the  robot.  Given  a
trajectory  of  the CoM  and  footstep  positions, inverse  kinematics
solvers can animate  the whole set of degrees of freemdom (DoFs) of the  
robot to generate a dynamically stable walk trajectory.

@@

This section presents an algorithm for constrained motion planning. 
The problem solved here
differs from classic approaches in two ways:
\begin{enumerate}
\item the set of valid configurations is defined implicitly, as the ones satisfying a given 
  set of inverse kinematics and balance tasks;
\item the goal is also defined implicitely, by additional inverse kinematic tasks.
\end{enumerate}
During global planning, we will consider several types of
tasks and constraints for various reasons:
\begin{itemize}
\item  Static stability: the center of mass of the robot stays at
  the vertical of the support polygon center, the two feet have fixed position and
  orientation.
\item End-effector position and orientation: the goals of some problems presented
  in the experimental section of this paper are defined as robot hand configuration,
  or gaze direction.
\item Configuration task: our adaptation of randomized motion planning algorithms 
  uses tasks defined as the distance towards a given configuration in $\mathcal{C}$. 
  This will be detailed in the following section.
\end{itemize}

@@

The equivalent random  extension on a constrained manifold $\mathcal{M}$, 
defined by the constraint function $f$,
starts from a valid configuration $q_{near} \in \mathcal{M}$, and extends the tree
towards a random configuration $q_{rand}$, while keeping the constraints
satisfied. Extension attempts orthogonal to $\mathcal{M}$ are useless,
as newly added edges have to be included  in $\mathcal{M}$. To extend in directions
that follow the directions of $\mathcal{M}$, we rely on the framework of 
jacobian-based prioritized inverse kinematics. Algorithm~\ref{alg:constrained} presents 
the adaptation of the classic extend function, and Fig.~\ref{fig:gikrrt} illustrates this
extension. The idea is to first project $q_{rand}$ on the tangent
space to $\mathcal{M}$  at $q_{near}$. Let us call the projected configuration $q_{rand}'$.
$q_{rand}'$ is obtained by one step of prioritized inverse kinematics optimization. Let 
$q_{rand}''$  be the result of a call to $\texttt{constraint-solver}(q_{rand}',f,\epsilon)$.  It is 
the projection
of $q_{rand}'$ on $\mathcal{M}$. Instead of extending the tree from $q_{near}$ towards 
$q_{rand}$, the algorithm tries to extend from $q_{near}$ towards 
$q_{rand}''$\footnote{This presentation attempts to give a precise idea of the algorithm, without 
  focusing on technical details. Readers interested in the actual implementation can
  refer to \url{https://github.com/laas/hpp-constrained} and 
  \url{https://github.com/laas/hpp-constrained-planner}, where the corresponding open-source 
  code is available.}. While extending the tree,
the configurations along the new edge are automatically projected onto $\mathcal{M}$. These projections 
are not very costly if the edge is close to the constrained manifold.

@@

The balanced walking conditions for a quasi-static walking robot are that the punctual
mass, or CoM, should always be over the support polygon (the convex hull of the two feet), and 
one foot can move iff the CoM is over the other foot. Similarly, the walking
conditions for a dynamic walking robot are that the ZMP should
always be in the robot support polygon, and one foot can move iff the ZMP is over the other 
foot. For a precise description of dynamically balanced walking conditions, the reader can refer
to \cite{wieber2002}. Under these assumptions, the following result holds:

@@

The model of a walking robot with a punctual
mass at a fixed height is known in the literature as the cart-table model \cite{kajita2003biped}.
The equations  giving  the   ZMP  horizontal  coordinates  $(p_x,p_y)$  as
functions  of CoM  horizontal coordinates $(x,y)$  in the  cart-table  model were
presented in \cite{kajita2003biped}:
\begin{equation}
\label{eq:walk-zmp}
\left(
\begin{array}{c}
p_x\\ p_y
\end{array}
\right) = \displaystyle \left(
\begin{array}{c}
x - \frac{z_c}{g} \ddot{x}\\ y - \frac{z_c}{g} \ddot{y}
\end{array}
\right)
\end{equation}
where $z_c$ is  the constant height of the CoM and  $g$ is the gravity
constant.    In    the    following    we    will    note    $\omega_0
=\sqrt{\frac{g}{z_c}}$.

@@

\subsubsection{Remarks}
\paragraph{Generalization to a complete model:} More work is needed to generalize 
the previous proof to any legged robot model. First, because we use a model where
the mass is punctual, we did not have to consider the angular-momentum of the robot
around the CoM, while it accounts for dynamic balance. Moreover, when considering
an actual kinematic and dynamic model,
the positions of feet and CoM are not controlled degrees of freedom, but inputs 
for inverse kinematic tasks. Continuity and regularity results about the robot
controller -- i.e. how much the configuration changes when the input of an inverse
kinematics task changes--, would be needed to generalize the proof. We have not
proven such properties for the inverse kinematics controller we use. 
The  proof presented above still gives the intuition for the
algorithm presented in the following.

@@

\caption{Solution path for a hand reaching problem in an
  appartment. The goal is implicitly defined as an inverse kinematics
  task.} 
\label{fig:shelf}
\end{figure}

@@

>>>>

In this paper, we present a planning algorithm that considers exact models of a humanoid
robot and its environment. It is used to solve navigation and manipulation problems. Our planner 
generates collision-free, statically stable 
geometric paths. We also present a constructive formal proof that these paths can be approximated by
collision-free, dynamically balanced walk trajectories. Our planner is designed for indoor
environments, where the floor is horizontal and flat. Because we do not compute explicitly footstep
positions at the planning stage, the presented planner is unable to plan motions in which the
robot steps over obstacles. This limitation is discussed in the paper.

@@

The problem of inverse kinematics for a humanoid robot, or any articulated
structure, is to compute a joint position to achieve an end-effector pose. As the
robots we deal with are redundant, it is natural to take advantage of
this redundancy by specifying multiple tasks, potentially with
different priorities. This problem has been widely studied in robotics
planning and control literature, and many Jacobian-based solutions have been
proposed, among which 
\cite{nakamura1986iks}, \cite{siciliano1991gfm},
\cite{baerlocher1998tpf} and \cite{khatib2004wbd}.
Obstacle avoidance can be taken into account with similar methods. To
do so, one has to include the obstacles as  constraints to
satisfy, see for example \cite{kanehiro2008lca}.
These methods are prone to fall into local minima, thus global motion
planning is needed to overcome this limitation.
\cite{TouGieGoe2007} propose a motion generation method where tasks follow
trajectories defined by cubic B-splines. The whole-body motion is optimized
with respect to the control point positions. This method can take into account
collisions with simple obstacles.

@@

When  planning a  whole-body motion  for a  humanoid robot, one difficult
challenge is to cope with  the curse of dimensionality. The complexity
of   motion  planning  is   exponential  in   the  dimension   of  the
configuration  space ($\mathcal{C}$)  to explore.  When  dealing with
high-dimensional configuration  spaces, it is  typically impossible to
explicitly represent  them, leading to the use  of randomized sampling
techniques  to solve  global planning  problems. In  the  past fifteen
years,  \textit{Probabilistic Roadmaps} \cite{kavraki1996prp} and  
\textit{Rapidly exploring Random  Trees} (RRT) 
\cite{kuffner00rrtconnect}  have been  developed and  used to  solve many
high-dimensional   planning  problems, see \cite{Lav06} and \cite{choset2005prm} for comprehensive
overviews.
When  using  sampling  techniques  on  a humanoid  robot,  another difficulty
is to  take into  account stability  constraints,  i.e. to
generate  random  configurations   on  zero  volume  submanifolds  of
$\mathcal{C}$. This problem has been investigated with success during
the last few years, \cite{Berenson15032011} presents a exhaustive survey
of Jacobian-based methods. Other recent contributions \cite{porta2012randomized}
present sophisticated constrained planning techniques based on higher-dimensional
continuation. Section \ref{sec:wb} presents a simple adaptation
of the RRT algorithm to constrained motion planning, that was first
introduced in \cite{dalibard09}.

@@

Another  field of  humanoid  robotics research  is  the generation  of
dynamically stable  walk patterns. Since  the introduction of  the ZMP
formalism  \cite{vukobratovic1969contribution},  several  methods  have  been  proposed  to
generate  walking  motions efficiently.   One  way  to  deal with  the
complexity  of a  humanoid  robotics  kinematic tree  is  to use  the
so-called "cart-table" simplified model \cite{kajita2003biped}. Based on this
model,  planning a trajectory  for the  ZMP is  reduced to  planning a
trajectory  for  the Center  of  Mass (CoM)  of  the  robot.  Given  a
trajectory  of  the CoM  and  footstep  positions, inverse  kinematics
solvers can animate  the whole set of degrees of freedom (DoFs) of the
robot to generate a dynamically stable walk trajectory.

@@

This section presents an algorithm for constrained motion planning. 
The problem solved here
differs from classic approaches in two ways:
\begin{enumerate}
\item the set of valid configurations is defined implicitly, as the ones satisfying a given 
  set of inverse kinematics and balance tasks;
\item the goal is also defined implicitly, by additional inverse kinematic tasks.
\end{enumerate}
During global planning, we will consider several types of
tasks and constraints for various reasons:
\begin{itemize}
\item Static stability: the center of mass of the robot stays above
  the support polygon center, the two feet have fixed position and
  orientation.
\item End-effector position and orientation: the goals of some problems presented
  in the experimental section of this paper are defined as robot hand configuration,
  or gaze direction.
\item Configuration task: our adaptation of randomized motion planning algorithms 
  uses tasks defined as the distance towards a given configuration in $\mathcal{C}$. 
  This will be detailed in the following section.
\end{itemize}

@@

The equivalent random  extension on a constrained manifold $\mathcal{M}$, 
defined by the constraint function $f$,
starts from a valid configuration $q_{near} \in \mathcal{M}$, and extends the tree
towards a random configuration $q_{rand}$, while keeping the constraints
satisfied. Extension attempts orthogonal to $\mathcal{M}$ are useless,
as newly added edges have to be included  in $\mathcal{M}$. To extend in directions
that follow the directions of $\mathcal{M}$, we rely on the framework of 
Jacobian-based prioritized inverse kinematics. Algorithm~\ref{alg:constrained} presents
the adaptation of the classic extend function, and Fig.~\ref{fig:gikrrt} illustrates this
extension. The idea is to first project $q_{rand}$ on the tangent
space to $\mathcal{M}$  at $q_{near}$. Let us call the projected configuration $q_{rand}'$.
$q_{rand}'$ is obtained by one step of prioritized inverse kinematics optimization. Let 
$q_{rand}''$  be the result of a call to $\texttt{constraint-solver}(q_{rand}',f,\epsilon)$.  It is 
the projection
of $q_{rand}'$ on $\mathcal{M}$. Instead of extending the tree from $q_{near}$ towards 
$q_{rand}$, the algorithm tries to extend from $q_{near}$ towards 
$q_{rand}''$\footnote{This presentation attempts to give a precise idea of the algorithm, without 
  focusing on technical details. Readers interested in the actual implementation can
  refer to \url{https://github.com/laas/hpp-constrained} and 
  \url{https://github.com/laas/hpp-constrained-planner}, where the corresponding open-source 
  code is available.}. While extending the tree,
the configurations along the new edge are automatically projected onto $\mathcal{M}$. These projections 
are not very costly if the edge is close to the constrained manifold.

@@

The balanced walking conditions for a quasi-static walking robot are that the point
mass, or CoM, should always be over the support polygon (the convex hull of the two feet), and 
one foot can move iff the CoM is over the other foot. Similarly, the walking
conditions for a dynamic walking robot are that the ZMP should
always be in the robot support polygon, and one foot can move iff the ZMP is over the other 
foot. For a precise description of dynamically balanced walking conditions, the reader can refer
to \cite{wieber2002}. Under these assumptions, the following result holds:

@@

The model of a walking robot with a point
mass at a fixed height is known in the literature as the cart-table model \cite{kajita2003biped}.
The equations  giving  the   ZMP  horizontal  coordinates  $(p_x,p_y)$  as
functions  of CoM  horizontal coordinates $(x,y)$  in the  cart-table  model were
presented in \cite{kajita2003biped}:
\begin{equation}
\label{eq:walk-zmp}
\left(
\begin{array}{c}
p_x\\ p_y
\end{array}
\right) = \displaystyle \left(
\begin{array}{c}
x - \frac{z_c}{g} \ddot{x}\\ y - \frac{z_c}{g} \ddot{y}
\end{array}
\right)
\end{equation}
where $z_c$ is  the constant height of the CoM and  $g$ is the gravity
constant.    In    the    following    we    will    note    $\omega_0
=\sqrt{\frac{g}{z_c}}$.

@@

\subsubsection{Remarks}
\paragraph{Generalization to a complete model:} More work is needed to generalize 
the previous proof to any legged robot model. First, because we use a model with
a point mass, we did not have to consider the angular-momentum of the robot
around the CoM, while it accounts for dynamic balance. Moreover, when considering
an actual kinematic and dynamic model,
the positions of feet and CoM are not controlled degrees of freedom, but inputs 
for inverse kinematic tasks. Continuity and regularity results about the robot
controller -- i.e. how much the configuration changes when the input of an inverse
kinematics task changes--, would be needed to generalize the proof. We have not
proven such properties for the inverse kinematics controller we use. 
The  proof presented above still gives the intuition for the
algorithm presented in the following.

@@

\caption{Solution path for a hand reaching problem in an
  apartment. The goal is implicitly defined as an inverse kinematics
  task.} 
\label{fig:shelf}
\end{figure}

@@

\section{Acknowledgments}

This work was supported by the French FUI Project ROMEO and the European Project ECHORD-231143. The authors would like to thank Thomas Moulard and Olivier Stasse for their valuable help during experiments.
